---
title: SDE Simulation
author: John Robin Inston
date: 2026-09-25
number-sections: true
title-block-banner: true
---

# SDE Simulation

In Financial Mathematics when pricing derivatives we often wish to simulate
continuous time stochastic price processes to in turn simulate derivative
payoffs and hence estimate option prices using **SDE Monte-Carlo Methods**. In
this note we will detail some of the methodologies employed to perform these
simulations as well as quantifying their error and work requirements.

Consider the general SDE with  differential form
$$
d X_{t} = f(t, X_{t})dt + g(t, X_{t})dW_{t},
$$
with Lipschitz $f,g$ and [[brownian-motion]] $W_{t}$. The integral form is the
exact solution
$$
X_{t}=X_{t_{0}}+\int_{t_{0}}^{t}{f(s, X_{s})}~d{s}+ \int_{t_{0}}^{t}{g(s,X_{s})}~d{W_{s}}.
$$
Supposing that no explicit solution exists we pose the problem of how can we
discretize and simulate the two integrals on the RHS of the expression?

We introduce two methodologies for simulating this process, first the simple
but flawed **Euler-Maruyama Method** followed by the more robust and commonly
used **Milstein Method**. We define the **weak** and **strong errors** of
convergence of SDE simulation and compute both error orders for each method. We
include several key examples including applications to option pricing, jump
processes, multiple correlated Brownian drivers and Brownian Bridges.

## Euler-Maruyama Method

The **Euler-Maruyama Method** is a SDE Simulation method with dynamics given by
stochastic differential equations. The [[ito-formula]] for a scalar-valued
function $U(t, X_{t})$ of the solution $X_{t}$ of the scalar Itô SDE has the
integral representation
$$
U(t,X_{t}) = U(t_{0}, X_{t_{0}})+ \int_{t_{0}}^{t}{L^0U(s,X_{s})}~d{s} + \int_{t_{0}}^{t}{L^1U(s, X_{s})}~d{W_{s}},
$$
where the differential operators $L^0$ and $L^1$ are defined by
$$
L^0 = \frac{\partial}{\partial t}+f \frac{\partial}{\partial x}+ \frac{1}{2}g^2 \frac{\partial^2}{\partial x^2}\quad \& \quad L^1 = g \frac{\partial}{\partial x}.
$$
This differs from the deterministic chain rule by the additional third term in
the $L^0$ operator, which is essentially due to the fact that
$\mathbb{E}[\Delta W^2]=\Delta t$. When $U(t,x)=x$ the Itô formula above is
simply
$$
X_{t}=X_{t_{0}}+\int_{t_{0}}^{t}{f(s,X_{s})}~d{s} + \int_{t_{0}}^{t}{g(s,X_{s})}~d{}W_{s}.
$$
Applying the Itô formula to $f$ and $g$ we obtain
$$
\begin{align}
X_{t} & =X_{0}  \\
 & \qquad+ \int_{t_{0}}^{t}{\left[ f(t_{0}X_{t_{0}})+\int_{t_{0}}^{s}{L^0f(u,X_{u})}~d{u}+\int_{t_{0}}^{s}{L^1 f(u, X_{u})}~d{W_{u}} \right]}~d{s}  \\
 & \qquad +\int_{t_{0}}^{t}{\left[ g(t_{0}, X_{t_{0}})+\int_{t_{0}}^{s}{L^0g(u, X_{u})}~d{u}+\int_{t_{0}}^{s}{L^1g(u, X_{u})}~d{W_{u}} \right]}~d{W_{s}} \\
 & =X_{t_{0}}+f(t_{0}, X_{t_{0}})\int_{t_{0}}^{t}{}~d{s}+g(t_{0}, X_{t_{0}})\int_{t_{0}}^{t}{}~d{W_{s}}+R_{1}(t,t_{0}),
\end{align}
$$
where the remainder is
$$
\begin{align}
R_{1}(t,t_{0}) & =\int_{t_{0}}^{s}{\int_{t_{0}}^{s}{L^0f(u, X_{u})}~d{u}}~d{s} + \int_{t_{0}}^{s}{\int_{t_{0}}^{s}{L^1f(u, X_{u})}~d{W_{u}}}~d{s} \\
 & + \int_{t_{0}}^{t}{\int_{t_{0}}^{s}{L^0g(u, X_{u})}~d{u}}~d{W_{s}} + \int_{t_{0}}^{t}{\int_{t_{0}}^{s}{L^1 g(u, X_{u})}~d{W_{u}}}~d{W_{s}}.
\end{align}
$$
Replacing $t_{0}$ by $t_{n}$, $t$ by $t_{n+1}$ and discarding the remainder be
obtain the **Euler-Maruyama Method** which we summarize in the definition
below. This is the simplest stochastic [[taylor-series|Taylor expansion]].

:::{.definition data-title="Euler-Maruyama Method"}

The **Euler-Maruyama Method** discretizes the continuous time dynamics by
defining a step-size $\Delta t=\frac{T}{N}$ for $N \in \mathbb{Z}$ and
computing the approximate solutions at times $t_{i}=i\Delta t$ using simple
left-endpoint Riemann sums in place of the integrals
$$
\widehat{X}_{t_{n+1}}=\widehat{X}_{t_{n}}+\Delta tf(t_{n},\widehat{X}_{t_{n}})+\underbrace{\Delta W_{n}}_{=\sqrt{\Delta t}\cdot\xi_{n}}g(t_{n},\widehat{X}_{t_{n}}),
$$
where $\Delta W_{n}=W(t_{n+1})-W(t_{n})\sim\mathcal{N}(0,\Delta t)$ (or
equivalently with $\xi_{n}\sim\mathcal{N}(0,1)$ for all $n$).

:::

**Example: BS Model EM Method and Error**
Consider the [[black-scholes-merton-model]] which consists of two assets, a
risk free asset $B$ and a stock $S$ with price dynamics given by
$$
\begin{align}dB_{t}&=rB_{t}dt \\dS_{t}& =\alpha S_{t}dt+\sigma S_{t}dW_{t},\end{align}
$$
on filtered probability space $(\Omega, \mathcal{F}, \mathbb{F}, \mathbb{P})$
where $r, \alpha$ and $\sigma$ are deterministic constants and $W_{t}$ is BM in
$\mathbb{P}$. Under the risk-neutral measure $\mathbb{Q}$ the dynamics of $S_t$
are instead given by
$$
dS_{t}=rS_{t}dt+\sigma S_{t}dW_{t}^\mathbb{Q},
$$
and the **true solution** under the risk-free measure is
$$
S_{T}=S_{0}e^{(r-\frac{\sigma^2}{2})T+\sigma W_{T}}.
$$
for some future time $T$. Suppose we wish to simulate this process on the
interval $[0,T]$. We can apply Euler's algorithm by considering a discretized
mesh $0=t_{0}, t_{1}, \dots, t_{n}=T$ where $t_{n}=t_{n-1}+\Delta t$ for
$\Delta t=\frac{T}{n}$ and then writing
$$
S_{t+\Delta t}=S_{t}+rS_{t}\Delta t + \sigma S_{t}\underbrace{(W_{t+\Delta t}-W_{t})}_{=\sqrt{ \Delta t }X_{t}},
$$
where $X_{t}\stackrel{iid}\sim\mathcal{N}(0,1)$. This is easy to simulate, for
example the following `R` code using a `for` loop:

```
## Euler's Algorithm
bs_euler <- function(r, alpha, sigma, S0, t0=0, T, n=1000){
	dt = T/n
	S <- numeric(n)
	S[1] <- S0
	for(i in 2:n){
		S[i] <- S[i-1] + r*S[i-1]*dt + sigma*S[i-1]*sqrt(dt)*rnorm(1,0,1)
	}
	return(S)
}
```

We have clearly made some kind of discretization error. To determine how bad
this error is we turn back to the true solution and we consider the Taylor
expansion
$$
\begin{align}
S_{t+\Delta t} & =S_{t}e^{\left( r-\frac{1}{2}\sigma^2 \right)\Delta t+\sigma \sqrt{ \Delta t }\xi}  \\
 & =S_{t}\left\{ \underbrace{1+\left(r-\frac{1}{2}\sigma^2\right)\Delta t+\sigma \sqrt{ \Delta t }\xi}_{\text{Euler's Approximation}}+\frac{1}{2}\left[ \left( r- \frac{1}{2}\sigma^2\right)\Delta t+\sigma \sqrt{ \Delta t }\xi \right]^2 + \dots  \right\}  \\
 & =S_{t}\left\{ 1+\sigma(\Delta t)^{\frac{1}{2}}\xi+\Delta t \left[ \left( r- \frac{1}{2}\sigma^2 \right)+\frac{1}{2}\sigma^2\xi^2 \right]+(\Delta t)^{\frac{3}{2}}\left[ \dots \right]+\dots   \right\},
\end{align}
$$
where we have replaced $W_{t+\Delta t}-W_{t}\sim \sqrt{ \Delta t }\xi$ where
$\xi\sim\mathcal{N}(0,1)$. From this we see that Euler's approximation might
seem reasonable from the SDE perspective but it is actually rather strange in
its use of various orders of $\Delta t$.  

It seems reasonable to suggest that the approximation improves as
$N \to \infty$. To test this presumption, we consider the two standard
approaches for measuring the error $X_{n}\to X(t_{n})$, leading to the concepts
of **weak error** and **strong error**.  

## Milstein Method

The **Milstein Method** applies a similar methodology to the
[[euler-maruyama-method]] but provides a greater degree of accuracy as it also
includes additional terms from the remainder in the stochastic Taylor expansion
above. We apply the [[ito-formula]] to the integrand $L^1g$ in the fourth
double integral of the remainder $R_{1}(t, t_{0})$ we obtain the stochastic
Taylor expansion
$$
\begin{align}
X_{t}=X_{0}+f(t_{0}, X_{t_{0}})\int_{t_{0}}^{t}{}~d{s}+g(t_{0}, X_{t_{0}})\int_{t_{0}}^{t}{}~d{W_{s}}+L^1 & g(t_{0}, X_{t_{0}})\int_{{t_{0}}}^{t}{\int_{t_{0}}^{s}{}~d{W_{u}}}~d{W_{s}}+R_{2}(t, t_{0}),
\end{align}
$$
where the remainder is
$$
\begin{align}
R_{2}(t, t_{0}) & =\int_{t_{0}}^{s}{\int_{t_{0}}^{s}{L^0f(u, X_{u})}~d{u}}~d{s} + \int_{t_{0}}^{s}{\int_{t_{0}}^{s}{L^1f(u, X(u))}~d{W_{u}}}~d{s}+\int_{t_{0}}^{t}{\int_{t_{0}}^{s}{L^0g(u, X_{u})}~d{u}}~d{W_{s}}  \\
 & \quad+ \int_{t_{0}}^{t}{\int_{t_{0}}^{s}{\int_{t_{0}}^{u}{L^0L^1g(v, X_{v})}~d{v}}~d{W_{u}}}~d{W_{s}}+\int_{t_{0}}^{t}{\int_{t_{0}}^{s}{\int_{t_{0}}^{u}{L^1L^1g(v, X_{v})}~d{W_{v}}}~d{W_{u}}}~d{W_{s}}.
\end{align}
$$
Making the same replacements as above and once again discarding the remainder
we obtain the Milstein Method which we summarize in the definition below.

:::{.definition data-title="Milstein Method"}

The **Milstein method** is
$$
X_{n+1}=X_{n}+\Delta t f(t_{n}, X_{n})+\Delta W_{n}g(t_{n}, X_{n})+L^1g(t_{n}, X_{t_{n}})\int_{t_{n}}^{t_{n+1}}{\int_{t_{n}}^{s}{}~d{W_{u}}}~d{W_{s}},
$$
where the double integral is given by
$$
\int_{t_{n}}^{t_{n+1}}{\int_{t_{n}}^{t}~d{W_{s}}}{}~d{W_{t}}=\frac{1}{2}((\Delta W_{n})^2-\Delta t).
$$

:::

**Example: BM Model Milstein Algorithm**
An alternative approach could be to instead use the Milstein Algorithm 
$$
S_{t+\Delta t}=S_{t}\left\{ 1+\sigma(\Delta t)^{\frac{1}{2}}\xi+\Delta t \left[ r+ \frac{1}{2}\left( \sigma^2(\xi^2-1) \right) \right] \right\},\tag{6}
$$
which has improved on Euler's approximation by grouping terms based on powers
of $\Delta t$ and removed all terms with powers $>1$. This is again very easy
to simulate with a `for` loop:

```
## Milstein's Algorithm
S <- S0
for(i in 2:length(t)){
	xi <- rnorm(1,0,1)
	S[i] <- S[i-1]*(1+sigma*h^(1/2)*xi + h*(r+(1/2)*(sigma^2*(xi^2 - 1))))
}
```

## SDE Simulation Error

### Weak Error

Given a function $\Phi$ from some class (e.g. polynomials with degree
$d\leq k$) we define the weak error by
$$
e_{\Delta t}^\text{weak}:=\sup_{0\leq t_{n}\leq T}\lvert \mathbb{E}[\Phi(X_{n})]-\mathbb{E}[\Phi(X(t_{n}))] \rvert .
$$
The weak error measures how well the given methodology can approximate the mean
of $\Phi(X(t))$ (weak convergence measures the error of the means). We say that
the method **converges weakly** if, for any $\Phi$ in this class
$$
e_{\Delta t}^\text{weak}\to 0~~\text{as}~~\Delta t \to 0.
$$
Further, we say that a method has **weak order of convergence** $p$ if there
exists a constant $K$ and a step size level $\Delta t^*$ (both dependent on
$\Phi$) such that
$$
e_{\Delta t}^\text{weak}\leq K\Delta t^p;~~\forall~ 0 < \Delta t\leq \Delta t^*,
$$
more specifically, we take $p$ to be the largest value for which this holds.

:::{.theorem data-title="EM Weak Convergence"}

For a wide class of problems the EM method converges with weak order one.

:::

*Proof Outline:*
	Update from Higham and Kloeden §9.3.

It turns out that it is not necessary to compute exact Brownian paths to obtain
weak order equal to one. It is sufficient for the increments $\Delta W_{n}$ to
have the correct low-order moments, hence we may us the **weak EM** method
$$
X_{n+1}=X_{n}+\Delta tf(X_{n})+\overline{\Delta W_{n}} g(X_{n}),
$$
where $\overline{\Delta W_{n}}$ are i.i.d. uniform random variables taking
value $\pm \sqrt{ \Delta t }$.

### Strong Error

At each time $t_{n}$ the absolute value of the error
$\lvert X_{n}-X(t_{n}) \rvert$ is a random variable. The strong error is
defined as
$$
e_{\Delta t}^\text{strong}:=\sup_{0\leq t_{n}\leq T}\mathbb{E}[\lvert X_{n}-X(t_{n}) \rvert ],
$$
i.e. the path-wise error averaged over all paths. Further, we say that a method
converges strongly if
$$
e_{\Delta t}^\text{strong}\to 0 ~~\text{as}~~ \Delta t \to 0,
$$
and we say that a method has **strong order $p$** on a class of SDEs if there
exists $K$ and $\Delta t^*$ such that
$$
e_{\Delta t}^\text{strong}\leq K \Delta t^p;~~\forall~0<\Delta t \leq\Delta t^*.
$$

:::{.theorem data-title="EM Strong Convergence"}

In general the EM method has strong order $p=\frac{1}{2}$.

:::

*Proof:*
	Update from proof outline given in Chapter 9.

The definition of strong convergence has important implications for individual
paths, namely if we have that
$$
\mathbb{E}\lvert X_{n}-X(t_{n}) \rvert \leq K\Delta t^{1/2},
$$
then by the [[markov-inequality]] we have that
$$
\mathbb{P}(\lvert X_{n}-X(t_{n}) \rvert \geq \Delta t^{1/4})\leq K\Delta t^{1/4},
$$
or in other words, along any path the error with be small with probability
close to 1.  

## SDE Simulation Examples

### Multiple Black-Scholes SDE Simulation

Suppose we wish to use the Euler scheme to simulate the following system of
SDEs
$$
\begin{cases}
dS_{t}^1 = rS_{t}^1dt+\sigma_{1}S_{t}^1dW_{t}^1 \\
dS_{t}^2=rS_{t}^2dt+\sigma_{2}S_{t}^2dW_{t}^2,
\end{cases}
$$
where we have that $d\left< W^1, W^1 \right>_{t}=\rho~dt$. For discretized time
step $\Delta t$ we have
$$
\begin{cases}
\check{S}_{n+1}^1=\check{S}_{n}^1+r\check{S}_{n}^1\Delta t+\sigma_{1}\check{S}_{n}^1\Delta W_{n+1}^1 \\
\check{S}_{n+1}^2=\check{S}_{n}^2+r\check{S}_{n}^2\Delta t+\sigma_{1}\check{S}_{n}^2 \underbrace{\Delta W_{n+1}^2}_{\rho\Delta W_{n+1}^1+\sqrt{ 1-\rho^2 }\Delta \tilde{W}_{n+1}}, \\
\end{cases}
$$
where $\tilde{W}_{n}\perp W_{n}^1$. The idea is that to avoid simulating the
correlated Weiner processes we instead represent $W_{n}^2$ in terms of
$W^1_{n}$ and some independent Weiner process $\tilde{W}_{n}$. If we write the
correlation matrix of $W_{n}^1$ and $W_{n}^2$ the
[[cholesky-decomposition|Cholesky decomposition theorem]] states there exists a
lower triangular matrix $A$ such that
$$
\begin{bmatrix}
1 & \rho \\
\rho & 1 
\end{bmatrix} = AA^T=\begin{bmatrix}
a_{11}  & 0 \\
a_{12} & a_{22}
\end{bmatrix} \begin{bmatrix}
a_{11} & a_{12} \\
0 & a_{22}
\end{bmatrix}=\begin{bmatrix}
a_{11}^2 & a_{11}a_{12} \\
a_{12}a_{11} & a_{11}^2 + a_{12}^2
\end{bmatrix},
$$
which gives that $a_{11}=1$, $a_{12}=\rho$ and $a_{22}=\sqrt{ 1-\rho^2 }$. Then
we have that
$$
\begin{bmatrix}
W^1 \\
W^2
\end{bmatrix} \stackrel{\mathcal{L}}{=}A\cdot \begin{bmatrix}
W^1  \\
\tilde{W}
\end{bmatrix},
$$
leading to our definition of our independent process and our new system
$$
\begin{cases}
\check{S}_{n+1}^1 = S_{n}^1 + r\check{S}_{n}^1\Delta t+\sigma_{1}\check{S}_{n}^1 \Delta W_{n+1}^1 \\
\check{S}_{n+1}^2=S_{n}^2+r\check{S}_{n}^2\Delta t +\sigma_{2}\check{S}_{n}^2\left( \rho \Delta W_{n+1}^1 + \sqrt{ 1-\rho^2 }\Delta \tilde{W}_{n+1} \right). 
\end{cases}
$$

### Heston Model SDE Simulation

A famous example of Multi-Dimenstional SDE simulation is the [[heston-model]]
for stochastic volatility whereby an asset price process $(S_{t})$ and
corresponding volatility process $(V_{t})$ have dynamics under the risk-neutral
measure given b
$$
\begin{cases}
dS_{t}=rS_{t}dt+S_{t}\sqrt{V_{t}}dW_{t}^1 \\
dV_{t}=\kappa(\theta-V_{t})dt+\eta \sqrt{ V_{t} }dW_{t}^2,
\end{cases}
$$
where $W^1, W^2$ are dependent Wiener processes with constant
$d\left< W^{1},W^2 \right>_{t}=\rho$. Suppose we wish to simulate these
processes using the Euler scheme. Then for discretized step $\Delta t$ we would
simulate
$$
\begin{cases}
\check{S}_{n+1} = \check{S}_{n} + r\check{S}_{n}^1\Delta t+\check{S}_{n}\sqrt{ \check{V}_{n}} \Delta W_{n+1}^1 \\
\check{V}_{n+1}=\check{V}_{n}+\kappa (\theta-\check{V}_{n})\Delta t +\eta \sqrt{\check{V}_{n}}\left( \rho \Delta W_{n+1}^1 + \sqrt{ 1-\rho^2 }\Delta \tilde{W}_{n+1} \right). 
\end{cases}
$$
We can define the following function in R that simulates stochastic volatility
via the Euler scheme:

```{r}
# Stochastic Volatility (Euler's Scheme)
stoch_vol_Euler <- function(S0, V0, t0, T, delta_t, r, kappa, theta, eta, rho) {
	time <- seq(t0, T, by=delta_t)
	S <- c(S0, rep(NA, T/delta_t))
	V <- c(V0, rep(NA, T/delta_t))
	W1 <- rnorm(T/delta_t, 0, delta_t)
	W2 <- rnorm(T/delta_t, 0, delta_t)
	for(n in 1:(length(time)-1)){
		V[n+1] = V[n] + kappa * (theta - V[n]) * delta_t + eta * sqrt(V[n]) * (rho * W1[n] + sqrt(1-rho^2)*W2[n])
		S[n+1] = S[n] + r * S[n] + S[n]*sqrt(V[n])*W1[n]
	}
	results <- data.frame(
		Price = S,
		Volatility = V	
	)
	return(results)
}
```

*We note that there are some subtle challenges to consider if we attempt the
same applying the Milstein scheme and this is something we covered in
assignment 1.*

### Jump Diffusion Processes SDE Simulation

The [[merton-jump-diffusion-model]] for stock price process $(S_{t})$ allows
the process to jump by incorporating a [[poisson-process|Poisson process]]
$(Nt)$ with rate $\lambda$ that is independent of the driving standard Weiner
process $(W_{t})$
$$
dS_{t}=rS_{t}dt+\sigma S_{t}dW_{t}+\eta S_{t}dN_{t}.
$$
First assume that the jump diffusion has **homogeneous rate**. To simulate such
a process we would:

1. Simulate the jump times; and then
2. Simulate the process (e.g. with Euler or Milstein) between the jump points.

With a Poisson process times between jumps are exponentially distributed and so
we can first simulate a collection of jump times that span over the time
interval using exponential random variables and `while` loops.

We then need to discretize the interval $[t_{0},T]$ with time step $\Delta t$
and proceed to apply Euler's method between jumps (accounting for the fact that
the jump points will intersect the discretization steps) and adding our jump
impact at the correct points.

Below we include an example function generating a jump diffusion process with
time homogeneous rate $\lambda$:

```{r}
# Jump Diffusion Process (Euler's Scheme)
jump_diff_Euler <- function(S0, t0, T, delta_t, r, sigma, eta, lambda){
	# Simulate jumps on the interval
	t <- rexp(1, lambda)
	jump_times <- c()
	while(t<T){
		jump_times <- append(jump_times, t)
		t <- t + rexp(1, lambda)
	}
	# Apply Eulers scheme between jumps
	times <- append(seq(t0, T, delta_t), jump_times) |> sort()
	S <- S0
	for(n in 1:(length(times)-1)){
		delta_times <- times[n+1] - times[n]
		jump_component <- ifelse(times[n] %in% jump_times, 1, 0)
		S[n+1] <- S[n] + r*S[n]*delta_times + sigma*S[n]*rnorm(1,0,delta_times) + eta*S[n]*jump_component
	}
	results <- data.frame(
		Time = times,
		Price = S
	)
	return(results)
}
```

Next we consider the case of a **process dependent rate** $\lambda(S_{t})$
which raises the question of whether we are still able pre-simulate the jump
times. The answer is yes if there is some upper bound
$$
\lambda(S_{t})\leq \overline{\lambda},
$$
through the use of [[poisson-process|thinned Poisson processes]]. The idea is
that we can simulate the jumps for this maximum rate $\overline\lambda$ and
then at each proposed jump point $j_{i}$ we can accept or reject the joint with
probabilities
$$
f(\tau_{i})=\begin{cases}
1 & \text{w.p.} \quad \frac{\lambda (S_{t})\tau_{i}}{\overline\lambda} \\
0 & \text{o.w.}
\end{cases}
$$
### Brownian Bridges SDE Simulation

The [[brownian-bridge]] is an SDE $(X_{t})$ characterized by the dynamics 
$$
\begin{cases}dX_{t}=\frac{b-X_{t}}{1-t}dt+dB_{t}\\ X_{0}=a,\end{cases}
$$
for $a,b \in \mathbb{R}$ and $0 \leq t < 1$. The solution to this SDE is given
by
$$
X_{t}=a(1-t)+bt+(1-t)\int_{0}^t \frac{1}{1-s}dB_{s}.
$$
Suppose we have some skeletal process (e.g. end of day prices) and we wish to
estimate intraday price movements assuming they have Brownian motion dynamics.
In this case we can simulate a Brownian bridge (e.g. using Euler's scheme)
$$
\check{X}_{n+1} = \check{X}_{n}+\frac{b-\check{X}_{n}}{1-n}\Delta t + \Delta W_{n+1}
$$
where we specify $\check{X}_0=a$, updating $a$ and $b$ for each new estimation
interval.