---
title: "PSTAT 10 Data Science Principles"
subtitle: "Lecture 11: Continuous Random Variables"
author:
  - name: "John Robin Inston"
    email: "johninston@ucsb.edu"
    orcid: "0009-0001-7285-6823"
    affiliation: "University of California, Santa Barbara"
date: today
date-format: "long"
bibliography: static/references.bib
format:
    revealjs:
        theme: custom.scss
        slide-number: c/t
        include-in-header: assets/theme-init.js
        include-after-body: assets/theme-toggle.js
        incremental: true
execute:
  echo: true
  warning: false
knitr:
  opts_chunk:
    fig.align: center
    dev: png
---

# Introduction

## 🔁 Review: Lecture 10

::: {.spacer-sm}
:::

👈 Last lecture we had a [very math heavy]{.accent} lecture covering:

- Random Variables (RVs)
- Probability Mass Functions (PMFs)
- Cumulative Distribution Functions (CDFs)
- Binomial Distribution
- `rbinom()`, `pbinom()` and `dbinom()` functions

## 👀 Outline: Lecture 11

::: {.spacer-sm}
:::

👇 This lecture is again a little math heavy, and we shall be looking at:

- Generalizing `rbinom()`, `pbinom()` and `dbinom()` functions
- Continuous RVs
- Continuous Uniform Distribution
- Normal (Gaussian) Distribution

# Recap

## 🔁 Recap — Discrete Random Variables

::: {.spacer-sm}
:::

::: {.fragment}

Recall the following definitions:

- A [random variable]{.accent} $X$ is a mathematical object taking numerical outputs corresponding to a random experiment.

:::

::: {.fragment}

::: {.spacer-sm}
:::

- A [probability mass function (PMF)]{.accent} of random variable $X$ is defined by
  $$
  p_X(k) = \mathbb{P}(X=k),
  $$
  which is non-zero for values $k$ in the support of $X$.

:::

::: {.fragment}

::: {.spacer-sm}
:::

- The [cumulative distribution function (CDF)]{.accent} of a random variable $X$ is defined by
  $$
  F_X(k) = \mathbb{P}(X\leq k).
  $$

:::

::: {.fragment}

::: {.spacer-sm}
:::

- The [expectation]{.accent} of (discrete) $X$ is defined by $\mathbb{E}[X] = \sum_{x\in\mathcal{X}} x\cdot\mathbb{P}(X=x).$

:::

## 🔁 Recap — Binomial Distribution

::: {.spacer-sm}
:::

::: {.fragment}

Consider a random variable $X$ that follows a $\text{Binomial}(n,p)$ distribution.

The [PMF]{.accent} and [CDF]{.accent} of $X$ respectively are

$$
\begin{align}
p_X(k) & = \mathbb{P}(X=k) = {n \choose k} \cdot p^k \cdot (1-p)^{n-k} \\
\\
F_X(k) & = \mathbb{P}(X\leq k) = \mathbb{P}(X=0) + \mathbb{P}(X=1) + ... +\mathbb{P}(X=k)
\end{align}
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

We have a quick practice with the `rbinom()`, `pbinom()` and `dbinom()` functions.

:::

## 🔁 Recap — `dbinom()`, `pbinom()` & `rbinom()`

::: {.spacer-sm}
:::

::: {.fragment}

- `dbinom()` determines the probability of $X$ taking a specific value, i.e. $\mathbb{P}(X=k)$.
- `pbinom()` determines the CDF of $X$ at a certain value, i.e. $\mathbb{P}(X\leq k)$.
- `rbinom()` generates realizations from a $X\sim\text{Binomial}(n,p)$ random variable.

:::

::: {.fragment}

::: {.spacer-sm}
:::

[**Examples:**]{.accent} for $X\sim\text{Binomial}(n=15, p=0.69)$,

```{r}
#| layout-ncol: 3
# P(X = 7)
dbinom(x = 7, size = 15, prob = 0.69)
# P(X <= 7)
pbinom(q = 7, size = 15, prob = 0.69)
# 5 realizations of X
rbinom(n = 5, size = 15, prob = 0.69)
```

:::

## 📋 General Functions

::: {.spacer-sm}
:::

::: {.fragment}

These functions each take the [general form]{.accent} of the base R functions that exist for the [key named distributions]{.accent} (some of which we will cover later).

:::

::: {.fragment}

::: {.spacer-sm}
:::

We summarize these functions in the table below:

| Binomial | Poisson | Geometric | Role |
|----------|---------|-----------|------|
| `dbinom()` | `dpois()` | `dgeom()` | PMF |
| `pbinom()` | `ppois()` | `pgeom()` | CDF |
| `qbinom()` | `qpois()` | `qgeom()` | Quantile |
| `rbinom()` | `rpois()` | `rgeom()` | Simulation |

: R Base Discrete Distribution Functions

::: {.spacer-sm}
:::

:::{.emphasize}
📌 Note that the discrete uniform is different since we use the `sample()` function instead.
:::

::: {.spacer-sm}
:::

:::

# Poisson Distribution

## 🐦 Poisson Distribution

::: {.spacer-sm}
:::

::: {.fragment}

::: definition

A [Poisson random variable]{.accent} describes the outcomes of an experiment [counting occurrences in a fixed time interval]{.accent} when the [mean rate of occurrence]{.accent} is known.

:::

:::

::: {.fragment}

::: {.spacer-sm}
:::

- E.g. the number of [emails received per hour]{.accent}, [customers arriving]{.accent} at a store per day, or [typos]{.accent} on a printed page.
- E.g. the number of [earthquakes]{.accent} in a region per year, or [radioactive decay events]{.accent} detected per second.

:::

::: {.fragment}

::: {.spacer-sm}
:::

Formally, the [PMF]{.accent} of the Poisson random variable $X\sim\mathcal{P}(\lambda)$ with mean rate $\lambda$ is:

$$
\mathbb{P}(X=k)=\frac{e^{-\lambda}\lambda^k}{k!};\quad k \in\{0,1,2,...\}.
$$

:::{.emphasize}
📌 The CDF of this distribution is quite complicated mathematically and is not covered in this course.
:::

:::

## 🐦 Poisson Distribution Plot

::: {.spacer-sm}
:::

::: {.fragment}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| out-height: "90%"
#| fig-cap: "Poisson distribution plot"
plot(0:20, dpois(0:20, 1), type = "n", ylab = "P(X=k)", xlab = "k", main = "Poisson Distribution Plot")
points(0:20, dpois(0:20, 1), pch = 20, col = "darkgreen")
lines(0:20, dpois(0:20, 1), col = "darkgreen")
points(0:20, dpois(0:20, 4), pch = 20, col = "red")
lines(0:20, dpois(0:20, 4), col = "red")
points(0:20, dpois(0:20, 10), pch = 20, col = "purple")
lines(0:20, dpois(0:20, 10), col = "purple")
legend(
    "topright",
    legend = c(
        expression(paste(lambda, "=1")),
        expression(paste(lambda, "=4")),
        expression(paste(lambda, "=10"))
        ),
    lty = 1,
    col = c("darkgreen", "red", "purple")
    )
```

:::

## 🐦 Poisson Distribution Functions

::: {.spacer-sm}
:::

::: {.fragment}

Recall the roles of the functions `dpois()`, `ppois()` and `rpois()`. Below we demonstrate how to use them.

[**Examples**]{.accent}

Compute $\mathbb{P}(X=3)$ when $X\sim\mathcal{P}(2)$.

```{r}
#| output-location: column
dpois(3, 2)
```

:::

::: {.fragment}

::: {.spacer-sm}
:::

Compute $\mathbb{P}(X>5)$ when $X\sim\mathcal{P}(3)$.

```{r}
#| output-location: column
1 - ppois(4, 3)
```

:::

::: {.fragment}

::: {.spacer-sm}
:::

Using 10000 simulations, estimate the mean of $X$ when $X\sim\mathcal{P}(3)$.

```{r}
#| output-location: column
rpois(10000, 3) |> mean()
```

:::

## 🎯 Example — Expectation of a Poisson RV

::: {.spacer-sm}
:::

::: {.fragment}

Since $\lambda$ is defined as the [mean rate of occurrence]{.accent}, the expectation of a Poisson random variable is simply

$$
\mathbb{E}[X] = \lambda.
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

Recall $X\sim\mathcal{P}(3)$ from our previous example. Applying the formula:

$$
\mathbb{E}[X] = 3.
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

This matches the simulation from the previous slide, which averaged to approximately 3:

```{r}
#| output-location: column
rpois(10000, 3) |> mean()
```

:::

::: {.fragment}

::: {.spacer-sm}
:::

:::{.emphasize}
📌 In general, for $X\sim\mathcal{P}(\lambda)$, $\mathbb{E}[X] = \lambda$.
:::

:::

# Continuous Random Variables

## 🔄 Motivation

::: {.spacer-sm}
:::

::: {.fragment}

So far we have been considering [discrete random variables]{.accent}, i.e. random variables with [discrete]{.accent} support:

- Bernoulli $\{0, 1\}$
- Binomial $\{0,1,...,n\}$
- Discrete Uniform $\{\text{item}_1, \text{item}_2, ..., \text{item}_k\}$
- Anything countable $\{..., -2, -1, 0, 1, 2, ...\}$

:::

::: {.fragment}

::: {.spacer-sm}
:::

What about [continuous random variables]{.accent} with [continuous]{.accent} support:

- Closed intervals $[0,1]$
- Open or partially open intervals $(0,1)$ and $[0,1)$
- Times, heights, weights, position on a dartboard or in a cube, etc.

:::

## ⚡ Probability Paradox

::: {.spacer-sm}
:::

::: {.fragment}

Remember that $\mathbb{P}(X=k)\neq 0$ for any value $k$ in the support of a [discrete]{.accent} random variable $X$.

::: {.spacer-sm}
:::

:::{.emphasize}
📌 [This is not true for a continuous random variable!]{.accent}
:::

::: {.spacer-sm}
:::

:::

::: {.fragment}

::: {.spacer-sm}
:::

- Recall we discussed that the probability of selecting a single number from a [continuous interval]{.accent} is zero.
- However, the probability for selecting a subinterval from a continuous interval is nonzero.

:::

::: {.fragment}

::: {.spacer-sm}
:::

Returning to our [dartboard]{.accent} example: the probability of hitting any single point is zero, but the probability of hitting a whole [region]{.accent} is not.

- E.g. the probability of hitting the [bottom half]{.accent} of the dartboard is [non-zero]{.accent}.
- In fact, with a [uniformly random]{.accent} throw, that probability is exactly $\frac{1}{2}$ — the region's share of the board's area.

:::

## 📊 Probability Density Functions

::: {.spacer-sm}
:::



::: {.fragment}

::: {.spacer-sm}
:::

- [Continuous random variables]{.accent} are described by density functions.
  - Similar to mass functions for discrete random variables but with no point masses.
  - Probabilities are no longer point masses but are instead described by [areas under curves]{.accent}.

:::

::: {.fragment}

::: {.spacer-sm}
:::

::: definition

The [probability density function (PDF)]{.accent} of a continuous random variable $X$ is a function $f_X(x)$ such that for any $a\leq b$,
$$
\mathbb{P}(a\leq X\leq b) = \int_a^b f_X(x)\,dx.
$$

:::

:::

::: {.fragment}

::: {.spacer-sm}
:::

::: definition

The [cumulative distribution function (CDF)]{.accent} of a continuous random variable $X$ is
$$
F_X(x) = \mathbb{P}(X\leq x) = \int_{-\infty}^x f_X(t)\,dt.
$$

:::

:::

# Continuous Uniform Distribution

## 🚌 Motivating Example

::: {.spacer-sm}
:::

::: {.fragment}

### Bus waiting times

To consider a practical example, suppose buses arrive at a stop every 12 minutes. If you arrive at the bus stop at a random time, how long will you wait?

- In the best case scenario?
- In the worst case scenario?

:::

::: {.fragment}

::: {.spacer-sm}
:::

Let $X$ be the time you wait for the bus, i.e. a [continuous random variable]{.accent} supported on $(0,12)$.

To [select a distribution]{.accent} we must consider what [assumptions]{.accent} to make.

:::

::: {.fragment}

::: {.spacer-sm}
:::

To make a claim about the [distribution]{.accent} of $X$ we make the following assumptions:

- We arrive at the bus stop at a [random time]{.accent}; and
- You are [equally likely to arrive at any point]{.accent} between the $n$-th and $(n+1)$-st bus.

:::

::: {.fragment}

::: {.spacer-sm}
:::

This is the continuous version of the uniform distribution, hence we define a [continuous uniform random variable]{.accent}.

:::

## 📏 Continuous Uniform Distribution

::: {.spacer-sm}
:::

::: {.fragment}

A [continuous uniform random variable]{.accent} $X\sim\mathcal{U}([a,b])$ with continuous support $[a,b]$ has [probability density function]{.accent}

$$
f_X(x) = \begin{cases}
\frac{1}{b-a} &\quad x\in[a,b] \\
0 & \quad \text{otherwise}.
\end{cases}
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

As a check, notice that the density integrates to 1 over its support, as required of any PDF:

$$
\int_a^b f_X(x)\,dx = \int_a^b \frac{1}{b-a}\,dx = \left[\frac{x}{b-a}\right]_a^b = \frac{b-a}{b-a} = 1.
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

In the case of our bus stop example the support is $[0,12]$, i.e. $a=0$ and $b=12$, and so for $X\sim\mathcal{U}([0,12])$ we have that

$$
f_X(x) = \frac{1}{12};\quad x\in[0,12].
$$

:::

## 📏 Visualizing the Uniform Density

::: {.spacer-sm}
:::

::: {.fragment}

```{r}
#| echo: false
#| fig-cap: "Probability Density Plot"
#| fig-width: 10
#| fig-height: 6
#| out-height: "90%"
plot(0:12, rep(1/12, 13), ylim = c(0,0.1), type = "n",
xlab = "Waiting Time", ylab = " ",
main = "Potential Waiting Times")
lines(0:12, rep(1/12, 13))
lines(c(0,0), c(0,1/12), lty = 3)
lines(c(12,12), c(0,1/12), lty = 3)
lines(c(-0.3,0), c(0,0))
lines(c(12,12.3), c(0,0))
```

:::

## 📈 Continuous Uniform CDF

::: {.spacer-sm}
:::

::: {.fragment}

You may have noticed that since we can only compute the probability that $X$ falls in an [interval]{.accent}, we are primarily interested in the [CDF]{.accent} of continuous random variables, since

$$
\mathbb{P}(p\leq X\leq q)=\mathbb{P}(X\leq q) - \mathbb{P}(X\leq p) = F_X(q) - F_X(p).
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

Often it is challenging to write a closed form expression for the CDF, but in the case of the continuous uniform we find that the CDF is given by:

$$
F_X(x) = \mathbb{P}(X\leq x) = \frac{x-a}{b-a};\quad x\in [a,b].
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

Using calculus, we can [derive this CDF]{.accent} directly from the PDF by computing:

$$
\begin{align}
F_X(x) & = \mathbb{P}(X\leq x) = \int_a^x f_X(y)\,dy \\
& = \int_a^x \frac{1}{b-a}\,dy = \left[\frac{y}{b-a}\right]_a^x \\
& = \frac{x-a}{b-a}.
\end{align}
$$

:::

## 🎯 Uniform Probabilities

::: {.spacer-sm}
:::

::: {.fragment}

Let's say we wish to compute [analytically]{.accent} the probability that the wait time is between 3 and 7 minutes. This is equivalent to computing the shaded area:

```{r}
#| echo: false
#| fig-cap: "Probability Density Plot"
#| fig-width: 10
#| fig-height: 6
#| out-width: "60%"
#| out-height: "90%"
plot(0:12, rep(1/12, 13), ylim = c(0,0.1), type = "n",
xlab = "Waiting Time", ylab = " ",
main = "Potential Waiting Times")
lines(0:12, rep(1/12, 13))
lines(c(0,0), c(0,1/12), lty = 3)
lines(c(12,12), c(0,1/12), lty = 3)
lines(c(-0.3,0), c(0,0))
lines(c(12,12.3), c(0,0))
rect(xleft=3, ybottom=0, xright=7, ytop=1/12, col="orange", border=NA)
lines(c(3,3), c(0,1/12), lty = 3)
lines(c(7,7), c(0,1/12), lty = 3)
```

:::

## 🧮 Uniform Integral

::: {.spacer-sm}
:::

::: {.fragment}

We can compute this quickly using the [CDF]{.accent}:

$$
\begin{align}
\mathbb{P}(3\leq X\leq 7) & = \mathbb{P}(X\leq 7) - \mathbb{P}(X\leq 3) \\
& = \frac{7-0}{12-0} - \frac{3-0}{12-0} = \frac{1}{3}.
\end{align}
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

This can clearly be computed using basic geometry, but to practice our calculus methods we can also arrive at the [same answer]{.accent} directly:

$$
\begin{align}
\mathbb{P}(3\leq X\leq 7) & = \int_3^7 f_X(x)dx \\
& = \int_3^7 \frac{1}{12}dx \\
& = \left[\frac{x}{12}\right]_3^7 \\
& = \frac{7}{12}-\frac{3}{12} = \frac{4}{12}=\frac{1}{3}.
\end{align}
$$

:::

## 🧮 `punif()` Function

::: {.spacer-sm}
:::

::: {.fragment}

R has the in-built function `punif()` for computing the [uniform CDF]{.accent} $\mathbb{P}(X\leq x)$, which has the syntax:

```{r}
#| eval: false
punif(q = x, min = a, max = b, lower.tail = TRUE)
```

- `q` is the quantile $x$
- `min` is the lower bound of the support $a$
- `max` is the upper bound of the support $b$

:::

::: {.fragment}

::: {.spacer-sm}
:::

[**Quick Examples:**]{.accent} for $X\sim\mathcal{U}([0,12])$,

```{r}
#| layout-ncol: 2
# P(X <= 6)
punif(6, 0, 12)
# P(X <= 12)
punif(12, 0, 12)
```

:::

::: {.fragment}

::: {.spacer-sm}
:::

Checking our computations earlier we have that:

```{r}
#| output-location: column-fragment
punif(7, 0, 12) - punif(3,0,12)
```

:::

## 📋 General Functions (Continuous)

::: {.spacer-sm}
:::

::: {.fragment}

Much like with discrete random variables we have 4 [standard functions]{.accent} for each distribution:

| Uniform     | Exponential | Normal    | Role       |
|-------------|-------------|-----------|------------|
| `dunif()`   | `dexp()`    | `dnorm()` | PDF        |
| `punif()`   | `pexp()`    | `pnorm()` | CDF        |
| `qunif()`   | `qexp()`    | `qnorm()` | Quantile   |
| `runif()`   | `rexp()`    | `rnorm()` | Simulation |

: R Base Continuous Distribution Functions

:::

::: {.fragment}

::: {.spacer-sm}
:::

:::{.emphasize}
📌 R has built-in `d`/`p`/`q`/`r` functions for [many more continuous distributions]{.accent} too — e.g. gamma (`dgamma()`), beta (`dbeta()`), chi-squared (`dchisq()`), Student's t (`dt()`), and log-normal (`dlnorm()`).
:::

:::

## 💪 Exercise — Continuous Uniform in R

::: {.spacer-sm}
:::

```{r}
#| echo: false
library(countdown)
countdown(minutes = 4)
```

Use the functions from the previous slide to answer the following questions about the continuous uniform random variable $X\sim\mathcal{U}(10,25)$:

::: {.nonincremental}
1. What is the probability that $X$ takes a value greater than 20? (Hint: You can set `lower.tail = FALSE`)
2. What is the probability that $X$ takes a value either greater than 19 or less than 15?
3. Use simulation to approximate the expected value of $X$.
:::

## ✅ Solution — Continuous Uniform in R

::: {.spacer-sm}
:::

::: {.fragment}

```{r}
#| layout-ncol: 3
punif(20, min = 10, max = 25, lower.tail = FALSE)
punif(19, min = 10, max = 25, lower.tail = FALSE) +
  punif(15, min = 10, max = 25, lower.tail = TRUE)
runif(10000, min = 10, max = 25) |>
    mean()
```

- `punif(20, ..., lower.tail = FALSE)` computes $\mathbb{P}(X>20)$ directly using the [upper tail]{.accent} of the CDF.
- Since the events $\{X>19\}$ and $\{X<15\}$ are [mutually exclusive]{.accent}, we can just [add]{.accent} their individual probabilities together.
- `runif(10000, ...)` simulates 10000 realizations of $X$, and `mean()` approximates $\mathbb{E}[X]$ via their [long-run average]{.accent}.

:::

# Normal Distribution

## 🔔 What is the Normal Distribution

::: {.spacer-sm}
:::

::: {.fragment}

The [normal (Gaussian) distribution]{.accent} is the most important distribution in statistics. It is more commonly known as the bell curve.

```{r}
#| echo: false
#| fig-align: "center"
#| fig-cap: "Distribution plot for a standard normal random variable."
#| fig-width: 10
#| fig-height: 5
#| out-height: "70%"
plot(
    seq(-4, 4, by = 0.01), dnorm(seq(-4, 4, by = 0.01)), type = "l",
    xlab = "x", ylab = expression(f(x)), main = "Normal Distribution (Bell Curve)"
)
```

:::

## 🔔 Normal Parameters

::: {.spacer-sm}
:::

::: {.fragment}

### What is the Normal Distribution?

- The [normal (Gaussian) distribution]{.accent} is the most important distribution in statistics.
- It is more commonly known as the [bell curve]{.accent} because of its symmetric, bell-shaped density.

:::

::: {.fragment}

::: {.spacer-sm}
:::

### Normal Random Variable

We denote a normal random variable $X\sim\mathcal{N}(\mu,\sigma^2)$.

:::

::: {.fragment}

::: {.spacer-sm}
:::

The normal distribution is described by [parameters]{.accent}:

- The [mean]{.accent} $\mu$ describes the location of the peak; and
- The [variance]{.accent} $\sigma^2$ describes the spread of the bell.

:::

::: {.fragment}

::: {.spacer-sm}
:::

In our previous plot above we used $\mu=0$ and $\sigma^2=\sigma=1$, which is known as a [standard normal distribution]{.accent}.

:::

## 🔔 Normal Parameter Plots

::: {.spacer-sm}
:::

::: {.fragment}

```{r}
#| echo: false
#| fig-align: "center"
#| fig-cap: "Multiple Normal Distributions"
#| fig-width: 10
#| fig-height: 5
#| out-height: "90%"
plot(
    NA, type = "n",
    xlab = "x", ylab = expression(f(x)), main = "Normal Distribution (Bell Curve)",
    xlim = c(-8,8), ylim = c(0,0.55)
)
lines(seq(-8, 8, by = 0.01), dnorm(seq(-8, 8, by = 0.01), mean = 0, sd = 1), col = "blue")
lines(seq(-8, 8, by = 0.01), dnorm(seq(-8, 8, by = 0.01), mean = 3, sd = 2), col = "darkgreen")
lines(seq(-8, 8, by = 0.01), dnorm(seq(-8, 8, by = 0.01), mean = 0, sd = 3), col = "red")
lines(seq(-8, 8, by = 0.01), dnorm(seq(-8, 8, by = 0.01), mean = -4, sd = 0.75), col = "orange")
legend("topright",
    c(expression(paste(mu, "=0, ", sigma, "=1")),
      expression(paste(mu, "=3, ", sigma, "=2")),
      expression(paste(mu, "=0, ", sigma, "=3")),
      expression(paste(mu, "=-4, ", sigma, "=0.5"))
    ),
    col = c("blue", "darkgreen", "red", "orange"),
    lty = 1
    )
```

:::

## 🔔 Normal PDF and CDF

::: {.spacer-sm}
:::

::: {.fragment}

### PDF

For normal random variable $X\sim\mathcal{N}(\mu,\sigma^2)$ the [density function]{.accent} is

$$
f_X(x) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(\frac{-(x-\mu)^2}{2\sigma^2}\right);\quad x\in\mathbb{R}.
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

### CDF

Hence if we wish to determine $\mathbb{P}(-1\leq X\leq 1)$ we determine

$$
\mathbb{P}(-1\leq X\leq 1) = \int_{-1}^1 \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(\frac{-(x-\mu)^2}{2\sigma^2}\right) dx,
$$

which isn't so easy to compute. Instead we can use R functions!

:::

## 🔔 Examples — Normal R Functions

::: {.spacer-sm}
:::

::: {.fragment}

Compute $\mathbb{P}(3\leq X\leq 10)$ when $X\sim\mathcal{N}(4,3^2)$.

```{r}
#| output-location: fragment
pnorm(10, mean = 4, sd = 3, lower.tail = TRUE) -
    pnorm(3, mean = 4, sd = 3, lower.tail = TRUE)
```

:::

::: {.fragment}

::: {.spacer-sm}
:::

Approximate the mean of the sum of two $X\sim\mathcal{N}(1,2^2)$ random variables using $n=10000$ simulations.

```{r}
#| output-location: fragment
(rnorm(10000, mean = 1, sd = 2) + rnorm(10000, mean = 1, sd = 2))  |>
    mean()
```

:::

::: {.fragment}

::: {.spacer-sm}
:::

Find the value $x$ such that $\mathbb{P}(X\leq x) = 0.9$ when $X\sim\mathcal{N}(5,2^2)$ — the [90th percentile]{.accent} — using `qnorm()`.

```{r}
#| output-location: fragment
qnorm(0.9, mean = 5, sd = 2)
```

:::

## 💪 Exercise — Normal Probability

::: {.spacer-sm}
:::

```{r}
#| echo: false
library(countdown)
countdown(minutes = 5)
```

Suppose the heights of students in class A are normally distributed with a mean $\mu=67$ inches and a standard deviation of $\sigma=3$ inches.

::: {.nonincremental}
1. What is the probability that a randomly selected student's height is between 63 and 70 inches?
:::

Now consider class B student heights, which we suppose are normally distributed with mean $\mu=70$ inches and a standard deviation of $\sigma=4$ inches.

::: {.nonincremental}
2. Use $n=10000$ simulations to determine the probability that a randomly selected student from class A is taller than a randomly selected student from class B, assuming the class heights are independent.
:::

## ✅ Solution — Normal Probability

::: {.spacer-sm}
:::

::: {.fragment}

```{r}
#| output-location: column
pnorm(70, mean = 67, sd = 3) - pnorm(63, mean = 67, sd = 3)
```

- We compute $\mathbb{P}(63\leq X\leq 70)$ for class A ($X\sim\mathcal{N}(67,3^2)$) as the [difference of two CDF values]{.accent}: $F_X(70)-F_X(63)$.

:::

::: {.fragment}

::: {.spacer-sm}
:::

```{r}
#| output-location: column
(rnorm(10000, mean = 67, sd = 3) > rnorm(10000, mean = 70, sd = 4)) |> mean()
```

- We simulate 10000 heights from [each class]{.accent} independently, then use `mean()` on the [logical comparison]{.accent} to approximate $\mathbb{P}(X>Y)$ — the proportion of simulations where the class A height exceeds the class B height.

:::

## 🤔 Interesting Result — Motivating Problem

::: {.spacer-sm}
:::

::: {.fragment}

Stepping outside the bounds of the course for a moment, we take a look at this last simulation.

Define $X\sim\mathcal{N}(67,3^2)$ and $Y\sim\mathcal{N}(70,4^2)$.

:::

::: {.fragment}

::: {.spacer-sm}
:::

We approximated via simulation:

$$
\mathbb{P}(X>Y)=\mathbb{P}(X-Y>0)
$$

But can we compute this [probability exactly]{.accent}?

:::

## 🤔 Interesting Result — Why is $X-Y$ Gaussian?

::: {.spacer-sm}
:::

::: {.fragment}

A remarkable fact from probability theory: any [linear combination]{.accent} of independent normal random variables is [itself normal]{.accent}.

Suppose $X_1\sim\mathcal{N}(\mu_1,\sigma_1^2)$ and $X_2\sim\mathcal{N}(\mu_2,\sigma_2^2)$ are independent. Then

$$
X_1 - X_2 \sim \mathcal{N}\left(\mu_1-\mu_2,\; \sigma_1^2+\sigma_2^2\right).
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

The [mean]{.accent} follows from linearity of expectation:

$$
\mathbb{E}[X_1-X_2] = \mathbb{E}[X_1] - \mathbb{E}[X_2] = \mu_1-\mu_2.
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

The [variance]{.accent} follows since $X_1$ and $X_2$ are independent:

$$
\text{Var}(X_1-X_2) = \text{Var}(X_1) + \text{Var}(-X_2) = \sigma_1^2+\sigma_2^2.
$$

:::

## 🤔 Interesting Result — Computing the Probability

::: {.spacer-sm}
:::

::: {.fragment}

Applying this to $X\sim\mathcal{N}(67,3^2)$ and $Y\sim\mathcal{N}(70,4^2)$ with $\mu_1=67,\sigma_1=3,\mu_2=70,\sigma_2=4$, we get

$$
X-Y \sim \mathcal{N}\left(67-70,\; 3^2+4^2\right) = \mathcal{N}(-3, 25) = \mathcal{N}(-3, 5^2).
$$

:::

::: {.fragment}

::: {.spacer-sm}
:::

So we can compute $\mathbb{P}(X>Y)=\mathbb{P}(X-Y>0)$ [explicitly]{.accent} as:

```{r}
#| output-location: column-fragment
pnorm(0, mean = -3, sd = 5, lower.tail = F)
```

:::

## 🤔 Interesting Result — Visualizing $X$, $Y$ and $X-Y$

::: {.spacer-sm}
:::

::: {.fragment}

```{r}
#| echo: false
#| fig-align: "center"
#| fig-cap: "Distributions of X, Y and X - Y"
#| fig-width: 10
#| fig-height: 5
#| out-height: "90%"
plot(
    NA, type = "n",
    xlab = "value", ylab = expression(f(x)), main = "X, Y and X - Y",
    xlim = c(-25, 85), ylim = c(0, 0.14)
)
lines(seq(-25, 85, by = 0.05), dnorm(seq(-25, 85, by = 0.05), mean = 67, sd = 3), col = "blue")
lines(seq(-25, 85, by = 0.05), dnorm(seq(-25, 85, by = 0.05), mean = 70, sd = 4), col = "darkgreen")
lines(seq(-25, 85, by = 0.05), dnorm(seq(-25, 85, by = 0.05), mean = -3, sd = 5), col = "red")
legend("topright",
    c(expression(paste("X ~ N(67, ", 3^2, ")")),
      expression(paste("Y ~ N(70, ", 4^2, ")")),
      expression(paste("X - Y ~ N(-3, ", 5^2, ")"))
    ),
    col = c("blue", "darkgreen", "red"),
    lty = 1
    )
```

:::

# Summary

## ✅ Topics Covered

::: {.spacer-sm}
:::

::: {.fragment}

::: {.spacer-sm}
:::

🤔 Today we covered a lot of material, including:

- Poisson Distributions
- Generalizing R Functions
- Continuous Random Variables
- Continuous Uniform Distribution
- Normal (Gaussian) Distribution

:::

## 📅 Next Class

::: {.spacer-sm}
:::

::: {.fragment}

::: {.spacer-sm}
:::

🤩 Next class we will conclude our exploration of probability by looking into:

- More Advanced Simulations!

:::
