---
title: "PSTAT 10 Data Science Principles"
subtitle: "Lecture 8: Plotting"
author:
  - name: "John Robin Inston"
    email: "johninston@ucsb.edu"
    orcid: "0009-0001-7285-6823"
    affiliation: "University of California, Santa Barbara"
date: today
date-format: "long"
bibliography: static/references.bib
format:
    revealjs:
        theme: custom.scss
        slide-number: c/t
        include-in-header: assets/theme-init.js
        include-after-body: assets/theme-toggle.js
        incremental: true
execute:
  echo: true
  warning: false
knitr:
  opts_chunk:
    fig.align: center
    dev: png
---

```{r}
#| echo: false
library(tidyverse)
```

# Introduction

## 🔁 Review: Lecture 7

::: {.spacer-sm}
:::

👈 Last lecture we studied the following topics:

- [Tibble Objects]{.accent};
- [Tibble Manipulation]{.accent};
  - Selecting;
  - Filtering;
  - Mutating;
- [Reading Data]{.accent}
  - `.txt`;
  - `.csv`; and
  - `.xlsx`.

## 👀 Outline: Lecture 8

::: {.spacer-sm}
:::

👇 Today we will continue looking at [describing]{.accent} and [visualizing]{.accent} external data, specifically:

- Descriptive statistics
- Base R Plotting
- Probability
- Discrete and Continuous Random Variables

# Exploring Real Data

## 📄 `.csv` Files

::: {.fragment}

**File Types Recap**

Last lecture we looked at several different file types that data can be stored as, including:

- `.txt` files
- `.csv` files
- `.xlsx` files

:::

::: {.fragment}

:::{.spacer-sm}
:::

**Why `.csv`?**

In this course we shall be using `.csv` files almost exclusively, and so we typically will only need the `read.csv()` function to load real world data.

:::

::: {.fragment}

:::{.spacer-sm}
:::

**Loading `Michelson.csv`**

The `Michelson` data is a famous dataset measuring the speed of light. I keep my data in a `data` subfolder, hence the path name below.

```{r}
Michelson <- read.csv("data/Michelson.csv")
head(Michelson, 3)
```

:::

## 🌪️ Messy Data Problems

::: {.fragment}

**The Reality of Real Data**

Real data is rarely clean and well behaved. It is often [messy]{.accent} and [unintuitive]{.accent}, and it is our job as data scientists and statisticians to [understand]{.accent}, [interpret]{.accent} and [draw meaning]{.accent} from the mess.

:::

::: {.fragment}

::: {.spacer-sm}
:::

**Key Tools**

To accomplish this goal we begin with several key tools:

- Summary Statistics
- Graphical Tools
- Statistical Measures (Correlation, Standard Deviation)
- Prior Knowledge
- Qualitative Analysis (understanding where the data comes from)

:::

## 📖 Statistics Terminology

::: {.fragment}

**Population & Sample**

Data consists of information from [observations]{.accent}, [counts]{.accent}, [measurements]{.accent} or [responses]{.accent}.

- The [population]{.accent} is the complete collection of all observations.
- The [sample]{.accent} is a representative group selected from a population.

:::

::: {.fragment}

::: {.spacer-sm}
:::

**Parameters & Statistics**

- A [parameter]{.accent} describes a [population characteristic]{.accent}; a [statistic]{.accent} describes a [sample characteristic]{.accent}.
- [Descriptive]{.accent} statistics summarize data.
- [Inferential]{.accent} statistics [test hypotheses]{.accent} (PSTAT 126).

:::

## 📊 Summary Statistics

::: {.fragment}

[Summary statistics]{.accent} are the key statistics of a sample that we always generate to get an idea about the [shape]{.accent} and [location]{.accent} of the data.

**Key Statistics**

1. Minimum
2. Lower Quartile
3. Median
4. Mean
5. Upper Quartile
6. Maximum
7. Spread Measures (Standard Deviation, Variance)
8. Outliers
9. Variable Correlation

:::

## 💪 Exercise — Summary Statistics

```{r}
#| echo: false
library(countdown)
countdown(minutes = 2)
```

The `summary()` function provides the first 6 statistics. The correlation and standard deviation are given by the `cor()` and `sd()` functions respectively.

::: {.nonincremental}
1. Change the `Michelson` object to a tibble using the function `tibble()`.
2. Select the column `ds16` using the `select()` function and apply the function `summary()`.
:::

## ✅ Solution — Summary Statistics

::: {.fragment}

- [Convert]{.accent} `Michelson` to a tibble using `tibble()`.
- [Select]{.accent} the `ds16` column using `select()`.
- [Apply]{.accent} `summary()` to obtain the key summary statistics.

:::

:::{.fragment}

```{r}
#| output-location: column
Michelson <- Michelson |> tibble()

Michelson |>
  select(ds16) |>
  summary()
```

:::

## 📦 Boxplots

::: {.fragment}

[Boxplots]{.accent} are a graphical visualization of most of the information provided by the `summary()` function.

```{r}
#| fig-width: 8
#| fig-height: 4
#| fig-align: "center"
#| out-height: "60%"
boxplot(iris[,-5], col = 2:5, xlab="Length/Width",
        main="Boxplot of Iris Variables")
```

:::

## 📈 Descriptive Statistics

::: {.fragment}

[Descriptive statistics]{.accent} help to summarize key characteristics of the data such as:

- its [location]{.accent}
- its [spread]{.accent}
- if its distribution has one peak ([unimodal]{.accent}) or many peaks
- how sharp these peaks are ([kurtosis]{.accent})
- whether the data is [skewed]{.accent} one way or the other

:::

::: {.fragment}

::: {.spacer-sm}
:::

:::{.emphasize}
📌 In this course we only consider location and spread.
:::

:::

## ⚖️ Measures of Centrality & Spread

::: {.fragment}

**Measures of Centrality**

- [Arithmetic average]{.accent} or [mean]{.accent} — use the function `mean()`
- Data midpoint, a.k.a. the [median]{.accent} — use the function `median()`
- Most frequent value, a.k.a. the [mode]{.accent} — use the function `mode()`

:::

::: {.fragment}

::: {.spacer-sm}
:::

**Measures of Spread**

- [Range]{.accent} (difference between maximum and minimum) — use the function `range()`
- [Interquartile Range]{.accent} (difference between upper and lower quartile)
- [Variance]{.accent} and [Standard Deviation]{.accent} — use the functions `var()` and `sd()`

:::

## 🧮 Sample Mean & Standard Deviation

::: {.fragment}

**Sample Mean**

The [sample mean]{.accent} is just the arithmetic average. For a sample $(x_1, ..., x_n)$ the sample mean $\bar{x}$ is given by

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^nx_i=\frac{x_1 +x_2 + \cdots + x_n}{n}
$$

:::{.emphasize}
📌 This is only an estimate of the [(true) population mean]{.accent}, which we denote $\mu$.
:::

:::

::: {.fragment}

:::{.spacer-sm}
:::

**Sample Standard Deviation**

The [sample standard deviation]{.accent} $s$ is:

$$
s = \sqrt{\frac{1}{n-1}\sum_{i=1}^n(x_i-\bar{x})^2}
$$

:::{.emphasize}
📌 This estimates the [(true) population standard deviation]{.accent} $\sigma$ (recall $\sigma^2$ is the [variance]{.accent}).
:::

:::

## 🔗 Relation Measures

::: {.fragment}

:::{.emphasize}
📌 [Relation measures]{.accent} describe how [two variables change together]{.accent} — i.e. how does one variable change if we change the value of another?
:::

:::

::: {.fragment}

::: {.spacer-sm}
:::

**Intuition**

- Both [increase]{.accent} together → [positive relationship]{.accent} (e.g. height & weight).
- One [increases]{.accent} as the other [decreases]{.accent} → [negative relationship]{.accent} (e.g. price & demand).
- No consistent pattern → [unrelated]{.accent} (e.g. shoe size & IQ).

:::

::: {.fragment}

::: {.spacer-sm}
:::

**Formal Measures**

[Covariance]{.accent} measures the joint variability of two variables, and leads to [correlation]{.accent} — a measure of the strength of their linear relationship.

```{r}
cov(iris$Petal.Length, iris$Petal.Width)
cor(iris$Petal.Length, iris$Petal.Width)
```

:::

## 📈 Example — Correlation Plot

::: {.fragment}

```{r}
#| echo: false
#| fig-cap: "Scatter plot with line of best fit."
plot(iris$Petal.Length, iris$Petal.Width, pch = 20,
  col = "darkgreen", xlab = "Petal Length", ylab = "Petal Width",
  main = "Correlation between petal length and petal width.")
abline(lm(iris$Petal.Width ~ iris$Petal.Length), col = "red",
  lty = "dashed")
text(x = 5, y = 0.5, col = "red", labels = paste("Correlation:",
  cor(iris$Petal.Length, iris$Petal.Width) |> round(2)))
```

:::

## 🧮 Covariance & Correlation

::: {.fragment}

**Covariance**

The [covariance]{.accent} between two variables $x:= (x_1, ..., x_n)$ and $y:= (y_1, ..., y_n)$ is given by

$$
\text{Cov}(x,y)=\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})(y_i-\bar{y})
$$

:::{.emphasize}
📌 The [covariance can be negative]{.accent}, which indicates a negative linear relationship (one goes up means the other goes down).
:::

:::

::: {.fragment}

:::{.spacer-sm}
:::

**Correlation**

From the definition of covariance we naturally define [correlation]{.accent}, denoted by the Greek letter rho $\rho$, of two variables $x$ and $y$ as

$$
\rho_{XY} = \frac{\text{Cov}(x,y)}{s_x \cdot s_y}
$$

We can compute the correlation between two vectors in R using the function `cor()`.

:::

## 💪 Exercise — Computing Summary Statistics

```{r}
#| echo: false
library(countdown)
countdown(minutes = 5)
```

We shall be considering the `faithful.csv` dataset, which contains a list of waiting times between eruptions and the duration of eruptions for the Old Faithful geyser in Yellowstone National Park.

::: {.nonincremental}
1. Download the data from Canvas and save it to your working directory.
2. Load the data into R and save it to the object `faithful`. Convert this to a `tibble()` object.
3. Using the functions `mean()` and `sd()`, compute the mean of `eruptions` and the standard deviation of `waiting`.
4. Use the functions `cov()` and `cor()` to compute both the covariance and the correlation between `eruptions` and `waiting`.
:::

## ✅ Solution — Computing Summary Statistics

::: {.fragment}

- [Load]{.accent} `faithful.csv` and convert it to a tibble.
- [Compute]{.accent} the mean of `eruptions` and the standard deviation of `waiting` using `mean()` and `sd()`.
- [Compute]{.accent} the covariance and correlation between `eruptions` and `waiting` using `cov()` and `cor()`.

:::

:::{.fragment}

```{r}
#| layout-ncol: 4
# load data
faithful <- read.csv("data/faithful.csv") |> tibble()
# mean
faithful$eruptions |> mean()
# standard deviation
faithful$waiting |> sd()
# covariance
cov(faithful$eruptions, faithful$waiting)
# correlation
cor(faithful$eruptions, faithful$waiting)
```

:::

## ⚠️ Problems with Descriptive Statistics

::: {.fragment}

**The Problem**

It is very important to practice [caution]{.accent} when using summary and descriptive statistics. They are very [simplistic]{.accent} and are easily fooled by more complex data patterns.

:::

::: {.fragment}

::: {.spacer-sm}
:::

**Example Data**

To demonstrate some of the problems you may run into we consider `problem.csv`, which can be found on Canvas.

```{r}
# load problem data
problem <- read.csv("data/problem.csv") |>
  tibble()
# inspect data
problem |> head(8)
```

:::

## ⚠️ Interpreting the Summary

::: {.fragment}

Applying `summary()` to `problem` we obtain the following summary statistics:

```{r}
#| output-location: column
summary(problem)
```

:::

::: {.fragment}

::: {.spacer-sm}
:::

We might make the following quick observations:

- The variable `x` is clustered around 48.
- The variable `y` is clustered around 405.

:::

:::{.fragment}

:::{.spacer-sm}
:::

```{r}
cor(problem$a, problem$b)
```

:::{.spacer-sm}
:::

Furthermore we might conclude that there is a significant linear relationship between variables `a` and `b`, as their correlation is `r round(cor(problem$a, problem$b), 2)`.

:::

## ⚠️ A Closer Look — Histogram

A closer look at `x` and `y`:

```{r}
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
par(mfrow = c(1,2))
hist(problem$x, breaks = 30)
hist(problem$y, breaks = 30)
```



## ⚠️ A Closer Look — Scatter Plot 

The relationship between `a` and `b` does not appear to be linear!

```{r}
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 6
#| out-width: "60%"
#| echo: false
plot(problem$a, problem$b, pch = 20, col = "darkgreen", xlab = "a", ylab = "b",
main="Scatter plot with line of best fit.")
abline(lm(problem$b ~ problem$a), col = "red", lty = "dashed")
```

# Plotting in Base R

## 🖼️ Types of Plots

::: {.fragment}

There are a wide variety of plots which all have their uses for data visualization and interpretation, including:

:::

::: {.columns}

::: {.column width="45%"}

::: {.fragment}

- Scatterplots
- Boxplots
- Histograms
- Barcharts
- Pie Charts
- Q-Q plots

:::

:::

::: {.column width="55%"}

::: {.fragment}

```{r}
#| fig-width: 15
#| fig-height: 8
#| out-width: "100%"
#| fig-align: "center"
#| echo: false
par(mfrow = c(2, 2))
boxplot(iris$Sepal.Length, main = "Boxplot", col = "darkgreen")
hist(iris$Sepal.Length, main = "Histogram", col = "darkgreen")
plot(iris$Sepal.Length, iris$Sepal.Width,
  main = "Scatterplot", pch = 20, col = "darkgreen")
barplot(table(iris$Species), main = "Barchart", col = "darkgreen")
```

:::

:::

:::

::: {.fragment}

**Example Data**

For an example we consider the `airquality` data in the `datasets` library.

```{r}
#| output-location: column
# load data
airquality <- datasets::airquality |>
  tibble()
airquality |> head(3)
```

:::


## 😬 An "Unprofessional" Plot


::: {.fragment}

:::{.spacer-sm}
:::

**A Basic Scatter Plot**

We produce a [scatter plot]{.accent} of `Wind` against `Temp` using the `plot()` function.

```{r}
#| fig-align: "center"
#| fig-cap: "A very basic scatter plot!"
#| fig-width: 12
#| fig-height: 6
#| out-width: "60%"
plot(airquality$Wind, airquality$Temp)
```

- The axes are not labelled and there is no title.
- There is no formatting.

:::

## 🎨 A More Professional Plot

::: {.fragment}

This is fine but not very professional. Fortunately, `plot()` has a lot of arguments we can use to tune our plot.

```{r}
#| fig-align: "center"
#| fig-width: 12
#| fig-height: 6
#| out-width: "60%"
#| fig-cap: "A more professional plot!"
plot(airquality$Wind, airquality$Temp,
  main = "Wind and Temperature in NYC, 1973",
  xlab = "Wind (mph)",
  ylab = "Temperature (degrees F)",
  pch = 20,
  col = "darkgreen",
  panel.first = grid()
)
```

:::

## 🎨 Further Improvements

::: {.fragment}

**What We Changed**

- Defined a plot [title]{.accent} using `main="..."`;
- Defined x and y [axis labels]{.accent} using `xlab="..."` and `ylab="..."`;
- Formatted our [point type]{.accent} to be solid `pch=20` and dark green `col="darkgreen"`.

:::

::: {.fragment}

:::{.spacer-sm}
:::

**Line of Best Fit**

We also might want to overlay a [line of best fit]{.accent} using the [linear model]{.accent} `lm()` function.

We will not cover the `lm()` function in this course, but in summary it fits a linear regression model between the `x` and `y` variables and we use this model to plot the line of best fit.

We add this line to our plot using the function `abline()`.

:::

## 📉 Line of Best Fit

::: {.fragment}

```{r}
#| output-location: slide
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "A more professional plot with a line of best fit!"
#| out-height: "90%"
# A more professional plot with best fit line
plot(airquality$Wind, airquality$Temp,
  main = "Wind and Temperature in NYC, 1973",
  xlab = "Wind (mph)",
  ylab = "Temperature (degrees F)",
  pch = 20,
  col = "darkgreen",
  panel.first = grid()
)
abline(lm(airquality$Temp ~ airquality$Wind), col = "red")
```

:::

## 📐 Formula Notation

::: {.fragment}

Notice in `lm()` we used R's [formula notation]{.accent} `y ~ x`. It is recommended to use formula notation for your plots as it makes your code clearer.

**Key Syntax**

- `y ~ x` reads as "[y modeled by x]{.accent}" — the [response]{.accent} goes on the left of `~`, [predictors]{.accent} go on the right.
- The `data = ...` argument lets you refer to columns [by name]{.accent} directly, without repeating `dataframe$` in front of every variable.
- Multiple predictors can be combined with `+`, e.g. `y ~ x1 + x2`.

:::

:::{.fragment}

```{r}
#| output-location: slide
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "A more professional plot with a line of best fit!"
#| out-height: "90%"
# A more professional plot with best fit line
plot(Temp ~ Wind, data = airquality,
  main = "Wind and Temperature in NYC, 1973",
  xlab = "Wind (mph)",
  ylab = "Temperature (degrees F)",
  pch = 20,
  col = "darkgreen"
)
grid()
abline(lm(airquality$Temp ~ airquality$Wind), col = "red")
```

:::

## 〰️ Plotting Continuous Functions

::: {.fragment}

**Discretizing**

- Computers think in [discrete terms]{.accent} and cannot represent [continuous time]{.accent}.
- To plot continuous functions, e.g. `sin()` or `exp()`, we [discretize]{.accent} them — redefining the interval as a sequence with [very fine discrete jumps]{.accent}.
- For example, on the interval $[0,10]$ we could define:

:::

:::{.fragment}

```{r}
#| output-location: column
x <- seq(0, 10, by = 0.01)
head(x)
```

:::

::: {.fragment}

:::{.spacer-sm}
:::

**Cosine Wave Example**

- We compute `cos(x)` and specify `type="l"`, which plots a [smooth line]{.accent} between the points rather than the points themselves.

:::

:::{.fragment}

```{r}
#| output-location: slide
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Cosine Wave"
#| out-height: "90%"
plot(x, cos(x),
  main = "Cosine Wave",
  xlab = "x",
  ylab = "cos(x)",
  col = "darkgreen",
  type = "l"
)
```

:::

## ➕ Overlaying Lines

::: {.fragment}

**Using `lines()`**

- We can [overlay several lines]{.accent} one on top of the other using the `lines()` function.
- `lines()` takes broadly the [same arguments]{.accent} as `plot()` (e.g. `col`, `type`, `lty`).

:::

::: {.fragment}

:::{.spacer-sm}
:::

**Adding a Legend**

- We add a [legend]{.accent} using the function `legend()`.
- The first argument (e.g. `"bottomleft"`) sets the [position]{.accent} of the legend.
- `legend = c(...)` gives the [labels]{.accent} for each line.
- `col = ...` and `lty = ...` should [match the plotted lines]{.accent} so the legend correctly identifies each one.

:::

::: {.fragment}

:::{.spacer-sm}
:::

```{r}
#| output-location: slide
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Cosine & Sin Waves"
#| out-height: "90%"
# A more professional plot with best fit line
plot(x, cos(x),
  main = "Cosine Wave",
  xlab = "x", ylab = "f(x)",
  col = "darkgreen", type = "l"
)
lines(x, sin(x), col = "red")
legend("bottomleft",
  legend = c("sin(x)", "cos(x)"),
  col = c("darkgreen", "red"), lty = 1)
```

:::

## 🔁 Iterating Function Plotting

::: {.fragment}

For a bit of fun we conclude by demonstrating how a `for` loop can be used to iteratively produce complicated plots:

- We first plot the [base cosine wave]{.accent}, then loop `i` from 1 to 100.
- Each iteration draws a [phase-shifted sine wave]{.accent} `sin(x + i/10)` using `lines()`.
- Setting `col = i` cycles through R's [default color palette]{.accent}, giving each new line a different color.
- The result is a dense fan of overlapping waves, all built up from a [single simple plotting call repeated]{.accent} inside the loop.

:::

:::{.fragment}

```{r}
#| output-location: slide
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Lots and lots of waves!"
#| out-height: "90%"
# A more professional plot with best fit line
plot(x, cos(x),
  main = "Cosine Waves",
  xlab = "x",
  ylab = "f(x)",
  col = "darkgreen",
  type = "l"
)
for(i in 1:100){
  lines(x, sin(x+(i/10)),
  col = i)
}
```

:::

## 💪 Exercise — Plotting in R

```{r}
#| echo: false
library(countdown)
countdown(minutes = 5)
```

Produce plots for $x^3$ and $e^x$ for `x` between 1 and 10.

::: {.nonincremental}
1. Define your `x` as a discretized sequence of values.
2. Define `x_cubed` as $x^3$ and `exp_x` as $e^x$.
3. Construct a plot of `x` against `x_cubed`, adjusting formatting as desired.
4. Add an additional line to the plot representing $e^x$.
5. Add an appropriate legend.
:::

## ✅ Solution — Plotting in R

::: {.fragment}

- [Discretize]{.accent} `x` between 1 and 10, then define `x_cubed` as $x^3$ and `x_exp` as $e^x$.
- [Plot]{.accent} `x` against `x_cubed` as a line, then [overlay]{.accent} `x_exp` using `lines()`.
- [Add]{.accent} a legend identifying each line.

:::

:::{.fragment}

```{r}
#| output-location: slide
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Exercise Plot"
#| out-height: "90%"
# define variables
x <- seq(1, 10, by = 0.01)
x_cubed <- x^3
x_exp <- exp(x)
# produce plot
plot(
  x, x_cubed,
  main = "Exercise Plot",
  xlab = "x", ylab = "f(x)",
  type = "l", col = "darkgreen"
)
lines(x, x_exp, col = "red", lty = 2)
legend(
  "topleft",
  legend = c("x^3", "exp(x)"),
  col = c("darkgreen", "red"), lty = 1:2
)
```

:::

# Summary

## ✅ Topics Covered

::: {.fragment}

::: {.spacer-sm}
:::

🤔 Today we looked into:

- Descriptive statistics
- Base R Plotting

:::

## 📅 Next Class

::: {.fragment}

::: {.spacer-sm}
:::

🤩 Next class we will temporarily leave behind data science topics as we study probability theory! The next four classes will be (roughly) split as follows:

1. Introduction to Probability
2. Discrete Random Variables
3. Continuous Random Variables
4. Monte Carlo Methods

:::
